Order reduction and partial MILP models \\ for certifying Robustness of DNNs globally.

### 2025-07-29 AAAI2026/global.tex
update proof
global

	\subsection{Classical encoding}

    The most straightforward idea to encode the global verification question is to 
    consider variable $x_j,\hat{x}_j$ and $x'_j,\hat{x}'_j$ for nodes of the image $I$ and of its perturbation $I'$,
    using the classical encoding (\ref{eq11}) for each variable, that we call $\mathcal{M}^{classical}$.

    use two identical MILP models, one for 
    , i.e., to build a model

    based on two identical MILP models $\mathcal{M},\mathcal{M}'$ with completely disjoint variables (and their own constraints) plus some extra constraints:
	\begin{enumerate}
		\item Add constraints for connecting input nodes $\mathcal{M},\mathcal{M}'$ to meet the requirement $|\boldsymbol{x}-\boldsymbol{y}| \leq \varepsilon$.
		\item Set the optimization objective as the difference between two variables of the same output node in two models $\mathcal{M},\mathcal{M}'$ (because we want to compute $\max|f(\boldsymbol{x}) -f(\boldsymbol{y}) |$).
	\end{enumerate}
	This large model contains twice as many binary variables as $\mathcal{M}$. The computational cost of solving an MILP model grows roughly exponentially with the number of binary variables, and hence it will cost much more time.

	We may relax some constraints (by changing binary variables to continuous variables) as in the case of local robustness. However, the problem is that even relaxing a few nodes can cause a significant loss of accuracy, which motivates us to explore other modeling approaches.

@@

	%	However, in practice, with a reasonable timeout, this simplified model can usually obtain a better bound.
	%	
	\subsection{Improving the Accuracy of the Simplified Model}
	We can remove the disadvantage and improve the accuracy of the above simplified model by modifying the constraints, at the cost of increased computational time — a trade-off between accuracy and speed.

	\begin{figure}[t!]
		\centering
	\hspace*{-10ex}
	\begin{tikzpicture}[scale=0.65]
		\begin{axis}[	axis on top, xlabel = \(x'_i\),
			ylabel = {\(y_i\)}, zlabel = \(\hat{y}_i\),
			set layers=default,
			xmax = 4, xmin = -4,
			ymax = 1, ymin = -1,		
			zmax = 1, zmin = -1,
			unit vector ratio=1 1 1, scale=2.5,  ytick   = {-1,0,1},
			yticklabels = {$-\gamma_i$,$0$,$\gamma_i$}, xtick = {0},
			xticklabels = {$0$}, ztick   = {-1,0,1},
			zticklabels = {$-\gamma_i$,$0$,$\gamma_i$},
			view={35}{14},
			]
			\addplot3[ fill=blue,opacity=0.1, fill opacity=0.4] 
			coordinates {
				(0,0,0) (-1,1,0) (-4,1,0) (-4,-1,0) (0,-1,0) (0,0,0)
			};

Given $\gamma_i$ as the upper bound of $y_i$, $\alpha_i,\beta_i$ be the upper and lower bound of $x_i,x_i'$, 
	the precise  constraints for $\hat{y}_i = \ReLU(x'_i+y_i)-\ReLU(x'_i)$ (along with $\hat{x}'_i=\ReLU(x'_i)$) are as follows:
	\begin{align*}
		& \begin{aligned}
			y_i + x'_i &\leq a\beta_i        &
			y_i &+ x'_i \geq (1-a)\alpha_i \\
			x'_i       &\leq a'\beta_i       & 
			x'_i       &\geq (1-a')\alpha_i \\
			\hat{y}_i  &\leq a\gamma_i       &
			\hat{y}_i  &\geq -a'\gamma_i \\
			\hat{y}_i  &\leq y_i + (1-a)\gamma_i  &
			\hat{y}_i  &\geq y_i - (1-a')\gamma_i \\
			\hat{y}_i  &\leq -x'_i + a\beta_i &
			\hat{y}_i  &\geq -x'_i + (1-a')\alpha_i \\
			\hat{y}_i  &\leq y_i + x'_i + (1-a)(-\alpha_i) &
			\hat{y}_i  &\geq y_i + x'_i + a'(-\beta_i)
		\end{aligned}
	\end{align*} Here, $a,a'$ are binary variables, and $a'$ is also the binary variable in the constraints for $\hat{x}'_i=\ReLU(x'_i)$.

>>>>

	\subsection{Extending the classical encoding}

    The most straightforward idea to encode the global verification question is to 
    consider variable $x_j,\hat{x}_j$ and $x'_j,\hat{x}'_j$ for nodes of the image $I$ and of its perturbation $I'$,
    using the classical encoding (\ref{eq11}) \cite{MILP} for each variable, 
    that we call $\mathcal{M}^{classical}$.
    Further:
    \begin{enumerate}
		\item For each input nodes $x_j,x'_j$ of $I,I'$, we add constraints
        $|x_j-x'_j| \leq \varepsilon$ 
        ($L_\infty$-perturbation - see next section for $L_1$).
		\item We set the optimization objective as maximize $x_i - x_j +x'_j - x'_i$ to compute $\beta^\varepsilon_{i,j}=\beta^\varepsilon_{j,i}$.
	\end{enumerate}

	This model contains twice as many (binary) variables as for local robustness. As the worst-case complexity of solving an MILP model 
    is exponential in the number of binary variables, it will be computationally intractable to compute $\beta^\varepsilon_{i,j}$ in this way, but for very small DNNs. As explained in the previous section, the standard way is to rely on relaxing at least some of the binary variables into linear variables.

    The problem is that the LP relaxation of this classical encoding is extremely inaccurate, as variables $x_j,x'_j$ dependencies are due solely on the function definition through common ancestor variables in previous layers - as this is approximated with LP, most of the dependencies will be lost.

    On top of being an issue with partial MILP models, LP relaxation is also used during the Branch and Bound process internal to MILP solvers to provide bounds for each branch. So even when all variables are opened and the MILP model is fully accurate, the classical encoding will likely not be efficient as well.

    We will check that in table \ref{table.classical} in Section Experimental Evaluation.

	\subsection{A novel "2v" model with {\em diff variables}}

    The idea to circumvent the loss of dependencies when applying linear relaxation is to switch the variables towards $x'_j,\hat{x}'_j, y_j, \hat{y}_j$, where $y_j, \hat{y}_j$ are the {\em diff variables}, with
    $y_j = x_j - x'_j$ and $\hat{y}_j= \hat{x}_j - \hat{x}'_j$. Intuitively, the diff variables will be small as they will be carrying the difference between both the image $I$ and its perturbation $I'$.
    We will use the classical MILP encoding for $\hat{x}'_j$.
    However, we cannot apply the classical MILP encoding (\ref{eq11}) \cite{MILP} on the diff variable $\hat{y}_j$. Instead we study the function of 
    $\hat{y}_j$ given $y_j$ and $x'_j$, and encode it into a novel MILP model.
    A first important observation is that $I,I'$ being perfectly symetrical,
    for $\gamma_i$ being a upper bound of $y_i$, then $-\gamma_i$ is a lower bound 
    for $y_i$.

	We first display on Figure \ref{fig.2v} the function $f(x'_i,y_i)=\hat{y}_i$, when we know that $y_i \in [- \gamma_i, \gamma_i]$. When $x'_i < -\gamma_i$, then $x_i <0$ and we have $\hat{y}_i = \ReLU(x_i) - \ReLU(x'_i) = 0-0=0$.
    For $x'_i > \gamma_i$, we have $x_i >0$ and $\hat{y}_i=y_i$. Between these two linear planes, we have two more linear planes presenting intermediate cases, so 4 linear planes in total.

	\begin{figure}[b!]
		\centering
	\hspace*{-5ex}
	\begin{tikzpicture}[scale=0.65]
		\begin{axis}[	axis on top, xlabel = \(x'_i\),
			ylabel = {\(y_i\)}, zlabel = \(\hat{y}_i\),
			set layers=default,
			xmax = 4, xmin = -4,
			ymax = 1, ymin = -1,		
			zmax = 1, zmin = -1,
			unit vector ratio=1 1 1, scale=2.5,  ytick   = {-1,0,1},
			yticklabels = {$-\gamma_i$,$0$,$\gamma_i$}, xtick = {0},
			xticklabels = {$0$}, ztick   = {-1,0,1},
			zticklabels = {$-\gamma_i$,$0$,$\gamma_i$},
			view={35}{14},
			]
			\addplot3[ fill=blue,opacity=0.1, fill opacity=0.4] 
			coordinates {
				(0,0,0) (-1,1,0) (-4,1,0) (-4,-1,0) (0,-1,0) (0,0,0)
			};

Assuming that $y_i \in [-\gamma_i, \gamma_i]$,
and that $x'_i \in [\alpha_i,\beta_i]$,
we have that $\hat{y}_i = \ReLU(x'_i+y_i)-\ReLU(x'_i)$ is the solution of:
	\begin{align*}
		& \begin{aligned}
			y_i + x'_i &\leq a\beta_i        &
			y_i &+ x'_i \geq (1-a)\alpha_i \\
			x'_i       &\leq a'\beta_i       & 
			x'_i       &\geq (1-a')\alpha_i \\
			\hat{y}_i  &\leq a\gamma_i       &
			\hat{y}_i  &\geq -a'\gamma_i \\
			\hat{y}_i  &\leq y_i + (1-a)\gamma_i  &
			\hat{y}_i  &\geq y_i - (1-a')\gamma_i \\
			\hat{y}_i  &\leq -x'_i + a\beta_i &
			\hat{y}_i  &\geq -x'_i + (1-a')\alpha_i \\
			\hat{y}_i  &\leq y_i + x'_i + (1-a)(-\alpha_i) &
			\hat{y}_i  &\geq y_i + x'_i + a'(-\beta_i)
		\end{aligned}
	\end{align*} Here, $a,a'$ are binary variables, and $a'$ is also the binary variable in the constraints for $\hat{x}'_i=\ReLU(x'_i)$.

@@

	%	However, in practice, with a reasonable timeout, this simplified model can usually obtain a better bound.
	%

### 2025-07-29 AAAI2026/notation.tex
global

    We use $\mathcal{M}$ to denote the whole MILP model, where each unstable ReLU are encoded in the above way with one integer variable. This is often too computationally intensive, as the worst-case complexity of MILP is exponential in the number of integer variables \cite{DivideAndSlide}.

@@

	Intermediate between these 2 extreme cases, there is {\em partial MILP} 
    (pMILP for short) to get trade-offs between accuracy and runtime:
	Let $X$ be a subset of the set of unstable neurons, and $n$ a neuron for which we want to compute upper and lower bounds on values: the pMILP based on $X$ to compute neuron $n$ uses the MILP encpoding (\ref{eq11}), where variable $a$ is:
	\begin{itemize}
		\item binary for neurons in $X$ (exact encoding of the ReLU),
		\item linear for neurons not in $X$ (linear relaxation).
	\end{itemize}
	We will denote the above model by MILP$_X$. We say that a node is {\em opened} if it is in $X$.

>>>>

    We use $\mathcal{M}$ to denote the whole MILP model, where each unstable ReLU are encoded in the above way with one integer variable. 
	The encoding from $(\hat{x}_j)_{j \text{ in layer } i}$ to 
	$(x_{j'})_{j' \text{ in layer } i+1}$ variables is simply the linear combination 
	$\boldsymbol{z}^{i} = \boldsymbol{W}^i\cdot \hat{\boldsymbol{z}}^{i-1}+ \vb^i$.
	This exact MILP encoding is often too computationally intensive, as the worst-case complexity of MILP is exponential in the number of integer variables \cite{DivideAndSlide}.

@@

	Intermediate between these 2 extreme cases, there is {\em partial MILP} 
    (pMILP for short) to get trade-offs between accuracy and runtime
	\cite{DivideAndSlide}:
	Let $X$ be a subset of the set of unstable neurons, and $n$ a neuron for which we want to compute upper and lower bounds on values: the pMILP based on $X$ to compute neuron $n$ uses the MILP encpoding (\ref{eq11}), where variable $a$ is:
	\begin{itemize}
		\item binary for neurons in $X$ (exact encoding of the ReLU),
		\item linear for neurons not in $X$ (linear relaxation).
	\end{itemize}
	We will denote the above model by MILP$_X$. We say that a node is {\em opened} if it is in $X$.
